A general Nullstellensatz for generalized spaces

### 2019-03-16 paper-qcoh.tex
Proof of the higher-order case

By \emph{geometric higher-order logic} we mean the extension of geometric
logic where we are allowed to form, in addition to the basic sorts supplied by
a given signature: finite limits of sorts; set-indexed colimits of sorts; and
powers of sorts. These derived sorts come with respective term constructors
(tuple formers, coprojections, set comprehension) and the usual rules governing
these constructors.

A \emph{geometric higher-order formula} is a formula built from equality and
relation symbols by the logical
connectives~${\top}\,{\bot}\,{\wedge}\,{\vee}\,{\exists}$
and by arbitrary set-indexed disjunctions~$\bigvee$. The existential
quantification can be over any sort, including the derived sorts. A
\emph{geometric higher-order sequent} is a sequent of the form~$(\varphi
\seq{\vec x} \psi)$ where~$\varphi$ and~$\psi$ are geometric higher-order
formulas and the sorts of the variables~$\vec x$ may be derived sorts.

The truth of geometric higher-order sequents is in general not preserved under
pullback along geometric morphisms; however for the fragment not containing
powersorts, this is true. In fact, there is a canonical way of associating to
any geometric higher-order sequent~$\sigma$ in this fragment a
family~$(\sigma_i)_i$ of ordinary geometric sequents such that~$\sigma$ is
provable in geometric higher-order logic if and only if all the
sequents~$\sigma_i$ are provable in ordinary geometric logic.

@@

\begin{thm}\label{thm:higher-order-nullstellensatz}
Let~$\TT$ be a geometric theory. Then, internally to~$\Set[\TT]$, for any
higher-order geometric$^\star$ sequent~$\sigma$ over the signature
of~$\ul{\TT}/U$, the following statements are equivalent:
\begin{enumerate}
\item The sequent~$\sigma$ holds for~$U_\TT$.
\item The sequent~$\sigma$ is provable$^\star$ modulo~$\TT/U_\TT$ in geometric
higher-order logic.
\end{enumerate}
\end{thm}

\begin{proof}The implication~$(2) \Rightarrow (1)$ is immediate since~$U_\TT$
is a model of~$\TT/U_\TT$. XXX
\end{proof}

>>>>

By \emph{extended geometric logic} we mean the extension of geometric logic
where we are allowed to form, in addition to the basic sorts supplied by
a given signature, finite limits of sorts and set-indexed colimits of sorts. By
(intuitionistic) \emph{higher-order logic}, we mean the further extension where
we may also form powersorts. These derived sorts come with respective term
constructors (tuple formers, coprojections, set comprehension) and the usual
rules governing these constructors.

An \emph{extended geometric formula} is a formula of extended geometric logic built from equality and
relation symbols by the logical
connectives~${\top}\,{\bot}\,{\wedge}\,{\vee}\,{\exists}$
and by arbitrary set-indexed disjunctions~$\bigvee$. The existential
quantification can be over any sorts of extended geometric logic, including the derived sorts. An
\emph{extended geometric sequent} is a sequent of the form~$(\varphi
\seq{\vec x} \psi)$ where~$\varphi$ and~$\psi$ are extended geometric
formulas and the sorts of the variables~$\vec x$ may be derived sorts.

It is possible to extend the Kripke--Joyal semantics so that higher-order logic
can be interpreted in any Grothendieck topos. The truth of a higher-order
sequent~$(\varphi \seq{\vec x} \psi)$ is in general not preserved under
pullback along geometric morphisms, even if~$\varphi$ and~$\psi$ do not
contain~${\forall}$ and~$\Rightarrow$, since powerobjects are in general not
preserved under pullback. However, as can be deduced from the following lemma,
the truth of extended geometric sequents is preserved; as is folklore, extended
geometric logic is just a thin layer over ordinary geometric logic.

\begin{lemma}\label{lemma:extended-conservative}
Let~$\sigma$ be an extended geometric sequent over the signature
of a geometric theory~$\TT$. Then there is a set-indexed family~$(\sigma_i)_i$ of ordinary
geometric sequents over the same signature, so that~$\sigma$ is provable in
extended geometric logic if and only if all the sequents~$\sigma_i$ are
provable in ordinary geometric logic.\end{lemma}

\begin{proof}Any existential quantification of the form~``$\exists p \? X
\times Y$'' can be replaced by the string~``$\exists x \? X\_ \exists y \?
Y$'', and similarly for free variables of product sorts appearing in the
context of~$\sigma$. In a similar vein more general finite limits are treated.

An existential quantification of the form~``$\exists x \? \coprod_i X_i$'' can
be replaced by the string~``$\bigvee_i \exists x \? X_i$''.

Finally, for any occurrence of a free variable~$x \?
\coprod_i X_i$ in the context of~$\sigma$, we can replace~$\sigma$ by the
family of sequents~$(\sigma_i)_i$, where the sequent~$\sigma_i$ is the same
as~$\sigma$ only that the free variable~$x$ is changed to be of sort~$X_i$ (and
the corresponding change in the consequent and the antecedent is applied,
applying the appropriate coprojection).

After carrying out these steps, the free variables are only of the basic sorts
supplied by the signature of~$\TT$ and existential quantifications only
range over the basic sorts. However, in the consequents and antecedents, still
tuple formers and coprojections may appear. These can be replaced as suggested
by the rules governing these. For instance, an equation~``$\langle x,y \rangle
= \langle x',y' \rangle$'' can be replaced by the conjunction~``$x = x' \wedge
y = y'$'', and an equation~``$\iota_i(x) = \iota_j(y)$'' (where~$\iota_i$
and~$\iota_j$ are coprojections associated with coproduct sorts) can be
replaced by the subsingleton-indexed disjunction~``$\bigvee\{ x = y \,|\, i = j \}$''.
\end{proof}

@@

\begin{thm}\label{thm:higher-order-nullstellensatz}
Let~$\TT$ be a geometric theory. Then, internally to~$\Set[\TT]$, for any
extended geometric$^\star$ sequent~$\sigma$ over the signature
of~$\ul{\TT}/U$, the following statements are equivalent:
\begin{enumerate}
\item The sequent~$\sigma$ holds for~$U_\TT$.
\item The sequent~$\sigma$ is provable$^\star$ modulo~$\TT/U_\TT$ in extended
geometric logic.
\end{enumerate}
\end{thm}

\begin{proof}The implication~$(2) \Rightarrow (1)$ is
immediate since~$U_\TT$ is a model of~$\TT/U_\TT$. The converse direction is by
Lemma~\ref{lemma:extended-conservative}, which holds internally in~$\Set[\TT]$
as the proof we supplied is constructive, and the Nullstellensatz for ordinary
geometric logic of Theorem~\ref{thm:nullstellensatz}.
\end{proof}
